---
format:
  gfm:
    preserve-yaml: false
    output-file: Monopsony_Robustness.md
    title-block-style: none
---
```{=markdown}
---
layout: default
title: Monopsony III — Further analysis
parent: Examples
nav_order: 36
math: true
description: "Varying the nuisance learner, the number of cross-fitting seeds, K, and cluster-vs-IID fold construction."
permalink: /examples/Monopsony_Robustness
enable_copy_code_button: true
---
```

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(knitr)
  library(modelsummary)
  library(stringr)
  library(ddml)
})

# ---------------------------------------------------------------------------
# Load per-seed intermediate results.
#
# Walks each subfolder of monopsony_data/intermediate (currently iid_K3,
# xclust_K3, xclust_K5), loads every cvc_seed*_K*.RData file via a fresh
# environment, and exposes one named list per subfolder (keyed by seed).
# Each list element contains the objects saved by Code/dube_monopsony/run_cvc.R
# (fit, fit_cvc, diag_obj, diag_cvc, fitted_obj).
# ---------------------------------------------------------------------------

intermediate_root <- '/Users/kahrens/PP Dropbox/Achim Ahrens/monopsony_data/dube_monopsony-selected/intermediate'

load_rdata_to_list <- function(file) {
  env <- new.env()
  load(file, envir = env)
  as.list(env)
}

for (sub in list.dirs(intermediate_root, recursive = FALSE)) {
  tag   <- basename(sub)
  files <- list.files(sub, pattern = "\\.RData$", full.names = TRUE)
  if (length(files) == 0L) next
  seeds <- as.integer(str_match(basename(files), "seed(\\d+)")[, 2])
  ord   <- order(seeds)
  files <- files[ord]; seeds <- seeds[ord]
  results <- lapply(files, load_rdata_to_list)
  names(results) <- sprintf("seed%d", seeds)
  assign(tag, results)
}

# Pulls the per-seed (or aggregated) cross-fitted R^2 for a given equation
# (y_X or D1_X) and learner index out of the diag_obj entries.
get_r2 <- function(fit_list, which = c("y_X", "D1_X"),
                   agg = "median", learner = 1) {
  which <- match.arg(which)
  out <- unlist(lapply(fit_list, function(x) x$diag_obj[[1]][[which]]$r2[learner]))
  if (identical(agg, "median")) return(median(out))
  if (identical(agg, "mean"))   return(mean(out))
  if (is.numeric(agg))          return(out[agg])
  stop("Invalid 'agg' argument. Use 'median', 'mean', or a seed index.")
}

# ---------------------------------------------------------------------------
# Helpers: convert ddml_plm summaries (per-seed or aggregated via ddml_rep)
# to modelsummary_list shims, so we can render with the same modelsummary +
# coef_map / gof_map style as the other two posts.
# ---------------------------------------------------------------------------

# Learner positions in the diagnostics r2 vector (matches the 12-learner
# order set in Code/dube_monopsony/run_cvc.R).
LEARNER_IDX <- c(
  "OLS" = 1L, "CV-Lasso" = 2L, "CV-Ridge" = 3L,
  "RF 1" = 4L, "RF 2" = 5L, "RF 3" = 6L,
  "XGB 1" = 7L, "XGB 2" = 8L, "XGB 3" = 9L,
  "NN 1" = 10L, "NN 2" = 11L, "NN 3" = 12L,
  "nnls1" = 13L, "singlebest" = 14L
)

# Build a modelsummary_list from a single ddml_plm fit (one seed) +
# its diagnostics object, picking a specific learner column.
ms_from_fit <- function(fit, diag_obj, learner = "XGB 3") {
  cm <- summary(fit)$coefficients[, , learner]
  est <- cm["D1", "Estimate"]
  se  <- cm["D1", "Std. Error"]
  zv  <- cm["D1", "z value"]
  pv  <- cm["D1", "Pr(>|z|)"]
  lidx <- LEARNER_IDX[[learner]]
  r2y <- diag_obj[[1]]$y_X$r2[lidx]
  r2d <- diag_obj[[1]]$D1_X$r2[lidx]
  out <- list(
    tidy = data.frame(
      term      = "log_reward",
      estimate  = unname(est),
      std.error = unname(se),
      statistic = unname(zv),
      p.value   = unname(pv)
    ),
    glance = data.frame(
      nobs = nobs(fit),
      r2_y = unname(r2y),
      r2_d = unname(r2d)
    )
  )
  class(out) <- "modelsummary_list"
  out
}

# Build a modelsummary_list from a seed list (e.g. xclust_K3) using
# ddml_rep median aggregation, plus the median-across-seeds R^2 from
# get_r2().
ms_from_rep <- function(seed_list, learner = "XGB 3") {
  fits   <- lapply(seed_list, function(x) x$fit)
  rep_o  <- ddml_rep(fits)
  cm     <- summary(rep_o)$coefficients[, , learner]
  est <- cm["D1", "Estimate"]; se <- cm["D1", "Std. Error"]
  zv  <- cm["D1", "z value"];  pv <- cm["D1", "Pr(>|z|)"]
  lidx <- LEARNER_IDX[[learner]]
  r2y <- get_r2(seed_list, which = "y_X",  agg = "median", learner = lidx)
  r2d <- get_r2(seed_list, which = "D1_X", agg = "median", learner = lidx)
  out <- list(
    tidy = data.frame(
      term      = "log_reward",
      estimate  = unname(est),
      std.error = unname(se),
      statistic = unname(zv),
      p.value   = unname(pv)
    ),
    glance = data.frame(
      nobs = nobs(fits[[1]]),
      r2_y = unname(r2y),
      r2_d = unname(r2d)
    )
  )
  class(out) <- "modelsummary_list"
  out
}

# Per-seed shim that pulls the seed-specific R^2 (agg = seed index) so the
# headline table reports honest per-seed predictive performance.
ms_from_seed <- function(seed_list, seed_idx, learner = "XGB 3") {
  entry <- seed_list[[seed_idx]]
  cm <- summary(entry$fit)$coefficients[, , learner]
  est <- cm["D1", "Estimate"]; se <- cm["D1", "Std. Error"]
  zv  <- cm["D1", "z value"];  pv <- cm["D1", "Pr(>|z|)"]
  lidx <- LEARNER_IDX[[learner]]
  r2y <- get_r2(seed_list, which = "y_X",  agg = seed_idx, learner = lidx)
  r2d <- get_r2(seed_list, which = "D1_X", agg = seed_idx, learner = lidx)
  out <- list(
    tidy = data.frame(
      term      = "log_reward",
      estimate  = unname(est),
      std.error = unname(se),
      statistic = unname(zv),
      p.value   = unname(pv)
    ),
    glance = data.frame(
      nobs = nobs(entry$fit),
      r2_y = unname(r2y),
      r2_d = unname(r2d)
    )
  )
  class(out) <- "modelsummary_list"
  out
}

gof_rows <- list(
  list(raw = "nobs", clean = "n",        fmt = 0),
  list(raw = "r2_y", clean = "R²(Y|X)",  fmt = 3),
  list(raw = "r2_d", clean = "R²(D|X)",  fmt = 3)
)
```

# Monopsony III — how robust is the headline elasticity?

*Part 3 of a three-post series. <a href="{{ '/examples/Monopsony_DML' | relative_url }}">Part 1</a> ran DDML on the Dube et al. (2020) MTurk data with hand-coded controls only. <a href="{{ '/examples/Monopsony_Finetune' | relative_url }}">Part 2</a> added fine-tuned DeBERTa embeddings. Both used $K = 3$, fold splitting by recruiter, a single XGBoost nuisance learner, one seed, and cluster-robust standard errors at the recruiter level. This post validates the results against other specification choices.*

This final part highlights validation checks worth running when applying DDML. We start by repeating the baseline cross-fitting under $S = 5$ different randomisation seeds and aggregating them with the paper's median-of-medians estimator. We then compare the resulting baseline median *pairwise* against three alternative variants — a different nuisance learner, IID instead of recruiter-honest fold construction, and a larger number of folds.

## Multi-seed aggregation

A single randomization seed generates one random partition of the sample into folds. The randomization seed also affects machine learning algorithms relying on randomization, such as random forests. Asymptotically the exact random partition does not matter; in finite samples it can make a difference. To avoid unnecessary reliance on randomizatin, we recommend repeating the cross-fitting $S$ times and reporting the median-aggregate estimate.

Below, we use the specification discussed in part Part 2: $K = 3$ recruiter-honest folds, XGBoost (the third XGB configuration, `XGB 3`) as the reported nuisance learner, and cluster-robust standard errors at the recruiter level. We then re-run this specification $S = 5$ times:

$$
\hat\theta_0^{\text{median}} = \mathrm{median}\bigl\{\hat\theta_{0,s}\bigr\}_{s=1}^{S},\qquad
\widehat{\mathrm{s.e.}}^{\text{median}} \;=\; \sqrt{\mathrm{median}\bigl\{\widehat{\mathrm{s.e.}}_s^{\,2} + (\hat\theta_{0,s} - \hat\theta_0^{\text{median}})^2\bigr\}_{s=1}^{S}}.
$$

```{r tbl-multiseed}
#| echo: false
#| results: asis
#| tbl-cap: "Per-seed estimates and median-of-medians aggregation for the baseline specification. XGB 3 nuisance learner; K = 3 recruiter-honest folds; cluster-robust SE per seed. R² rows are seed-specific for the per-seed columns and median-across-seeds for the aggregate column."
seed_names <- names(xclust_K3)
per_seed <- setNames(
  lapply(seq_along(xclust_K3),
         function(i) ms_from_seed(xclust_K3, i, learner = "XGB 3")),
  seed_names
)
tbl1 <- c(per_seed,
          list("median-of-medians (S=5)" = ms_from_rep(xclust_K3, "XGB 3")))
modelsummary(
  tbl1,
  coef_map  = c("log_reward" = "log(reward)"),
  gof_map   = gof_rows,
  fmt       = 3,
  statistic = "std.error",
  output    = "html"
)
```

The single-seed estimates span a non-trivial range relative to any individual seed's SE, so sample-split randomness has a noticeable impact in finite samples. We can be more confident in the median-of-medians estimate, which is less sensitive to outlier seeds.

### Choice of nuisance learner

The choice of nuisance function estimator is consequential for DDML estimation. Poorly chosen or poorly tuned leaners may result in misleading DDML point estimates. Below we compare our reference specification relying on `XGB 3` against CV-Lasso:

```{r tbl-learner}
#| echo: false
#| results: asis
#| tbl-cap: "Baseline (XGB 3) vs. CV-Lasso, both aggregated over S = 5 seeds. K = 3 recruiter-honest folds; cluster-robust SE."
tbl2 <- list(
  "XGB 3" = ms_from_rep(xclust_K3, learner = "XGB 3"),
  "CV-Lasso"         = ms_from_rep(xclust_K3, learner = "CV-Lasso"),
  "Stacking NNLS"    = ms_from_rep(xclust_K3, learner = "nnls1")
)
modelsummary(
  tbl2,
  coef_map  = c("log_reward" = "log(reward)"),
  gof_map   = gof_rows,
  fmt       = 3,
  statistic = "std.error",
  output    = "html"
)
```

The CV-Lasso underperforms badly on both $R^2(Y\mid X)$ and $R^2(D\mid X)$. The resulting estimate is indistinguishable from zero. 

In Section 6 , we discuss several options to constructing and validating nuisance function estimators. One straightforward approach is short-stacking, a combination of DDML and model averaging that leverages the cross-fitting where to choosing and validating nuisance function estimators is model averaging: rather than regress $Y$ and $D$ on the cross-fitted predicted values of every candidate learner under non-negative weights summing to one. Table 6 of the paper reports stacked DDML at $\hat\theta_0 = -0.054$ (s.e. $0.020$); the stacking weights load almost entirely on the three XGBoost specifications, so when XGBoost dominates the linear alternatives — as it does here — stacking inherits its performance for free.

```{r tbl-learner}
#| echo: false
#| results: asis
#| tbl-cap: "Baseline (XGB 3) vs. CV-Lasso, both aggregated over S = 5 seeds. K = 3 recruiter-honest folds; cluster-robust SE."
tbl2 <- list(
  "XGB 3" = ms_from_rep(xclust_K3, learner = "XGB 3"), 
  "Stacking NNLS"    = ms_from_rep(xclust_K3, learner = "nnls1")
)
modelsummary(
  tbl2,
  coef_map  = c("log_reward" = "log(reward)"),
  gof_map   = gof_rows,
  fmt       = 3,
  statistic = "std.error",
  output    = "html"
)
```

### 2.2 Cluster-respecting vs IID folds

```{r tbl-folds}
#| echo: false
#| results: asis
#| tbl-cap: "Baseline (recruiter-honest folds, cluster-robust SE) vs. IID folds with het-robust SE. Both columns are median-of-medians over S = 5 seeds; K = 3; XGB 3 nuisance learner."
tbl3 <- list(
  "xclust (baseline)" = ms_from_rep(xclust_K3, learner = "XGB 3"),
  "IID folds"         = ms_from_rep(iid_K3,    learner = "XGB 3"),
  "xclust (baseline)" = ms_from_rep(xclust_K3, learner = "XGB 3"),
  "IID folds"         = ms_from_rep(iid_K3,    learner = "XGB 3")
)
modelsummary(
  tbl3,
  coef_map  = c("log_reward" = "log(reward)"),
  gof_map   = gof_rows,
  fmt       = 3,
  statistic = "std.error",
  output    = "html"
)
```

This is the sweep where leaving the default behind is the point. The Ipeirotis cross-section has many postings per recruiter, and the same recruiter's HITs share unobserved features (recruiter style, payout norms, qualification preferences) that correlate with both posted reward and fill time. Splitting at the observation level lets siblings from the same recruiter sit on both sides of the fold boundary — making the residuals look more independent than they are, and the cross-fitted $R^2$ look better than the true generalisation $R^2$. In `ddml_plm` the lever is whether `cluster_variable` is supplied; passing the recruiter id forms recruiter-honest folds and clusters SEs at the same level; omitting it gives random IID folds and uses heteroskedasticity-robust SEs:

```r
fit <- ddml_plm(
  y = y, D = D, X = X,
  learners      = list(xgb_spec),
  sample_folds  = 3,
  ensemble_type = "singlebest",
  shortstack    = FALSE,
  cluster_variable = rid,   # omit for IID folds + het-robust SE
  silent        = TRUE
)
```

The cross-fitted $R^2$ values typically *fall* when fold construction moves from IID to recruiter-honest, because the held-out fold now contains recruiters the nuisance learner never saw — that is the honest predictive performance and it is what should enter the DDML score. The standard error typically *rises*, because clustering at the recruiter level accounts for the correlated residuals that het-robust ignores. The cluster-honest combination is the inference-bearing one for this application, which is why Entries 1–2 and the baseline of this post use it as their default.

### 2.3 Number of folds $K$

```{r tbl-K}
#| echo: false
#| results: asis
#| tbl-cap: "Baseline K = 3 vs. K = 5. Both columns are median-of-medians over S = 5 seeds; XGB 3 nuisance learner; recruiter-honest folds; cluster-robust SE."
tbl4 <- list(
  "K = 3 (baseline)" = ms_from_rep(xclust_K3, learner = "XGB 3"),
  "K = 5"            = ms_from_rep(xclust_K5, learner = "XGB 3")
)
modelsummary(
  tbl4,
  coef_map  = c("log_reward" = "log(reward)"),
  gof_map   = gof_rows,
  fmt       = 3,
  statistic = "std.error",
  output    = "html"
)
```

Cross-fitting theory accommodates any fixed $K$. Velez (2024) shows performance improves with more folds, peaking at $K = n$ with diminishing returns. In the paper we recommend the largest $K$ consistent with available compute, plus a sensitivity check. Moving from $K = 3$ to $K = 5$ trains each nuisance fit on $4/5$ of the data instead of $2/3$ and compute scales roughly linearly. With $n \approx 258{,}000$ either value is in the range where the bias–variance tradeoff is favourable.

## 3. What to take away

- A single ML learner is not a complete DDML specification. The point estimate is informative only when the nuisance learner can actually predict — cross-fitted $R^2$ is the diagnostic to lean on.
- Short-stacking (or, more conservatively, reporting both stacking and single-best) sidesteps the learner-selection problem and is cheap relative to refitting four nuisance models from scratch.
- Median-of-medians over $S \ge 3$ seeds is the inference-bearing point estimate; one-seed numbers are fine for iteration but should not be the published number.
- Cluster vs IID fold construction is the design decision with the largest swing in SE and the most consequential one for inference. The blog defaults to recruiter-honest folds + cluster-robust SEs; the IID variant is included only as a contrast.
- $K$ matters less than the other knobs in this application.

## References

- Ahrens, A., V. Chernozhukov, C. Hansen, D. Kozbur, M. Schaffer and T. Wiemann. *An Introduction to Double/Debiased Machine Learning.* §5 (median aggregation), §6 (Dube application, Tables 5–6), §7 (implementation guidance).
- Ahrens, A., C. B. Hansen, M. E. Schaffer and T. Wiemann (2024). [`ddml`: Double/debiased machine learning in R.](https://doi.org/10.18637/jss.v108.i03)
- Chernozhukov, V., D. Chetverikov, M. Demirer, E. Duflo, C. Hansen, W. Newey and J. Robins (2018). [Double/Debiased machine learning for treatment and structural parameters.](https://doi.org/10.1111/ectj.12097)
- Dube, A., J. Jacobs, S. Naidu and S. Suri (2020). [Monopsony in online labor markets.](https://doi.org/10.3386/w26108) *AER: Insights* 2 (1).
- Lei, J. (2020). Cross-validation with confidence. *Journal of the American Statistical Association.*
- Velez, A. (2024). On the asymptotic properties of debiased machine learning estimators.
