---
format:
  gfm:
    preserve-yaml: false
    output-file: AngristEvans.md
    title-block-style: none
---

```{=markdown}
---
layout: default
title: DDML for the Angrist & Evans
parent: Examples
nav_order: 40
math: true
description: "Imposing no-interaction restrictions in DML estimations."
permalink: /examples/AngristEvans
enable_copy_code_button: true
---
```

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(arrow)
library(knitr)

res <- read_parquet("results.parquet")

# Friendly learner labels. The single-learner names emitted by ddml are
# custom_1..custom_5, corresponding to the spec order in make_learner_specs():
# 1=OLS (base controls), 2=Lasso (poly dict), 3=Ridge (poly dict),
# 4=XGBoost lr=.01, 5=XGBoost lr=.03. "nnls" is the stacked ensemble.
# "ols" is the 2SLS row (estimator == "tsls").
learner_lbl <- c(
  "ols"      = "2SLS",
  "nnls"     = "Stacking",
  "custom_1" = "OLS",
  "custom_2" = "Lasso",
  "custom_3" = "Ridge",
  "custom_4" = "XGB(.01)",
  "custom_5" = "XGB(.03)"
)
learner_order <- c("2SLS", "Stacking", "OLS", "Lasso", "Ridge", "XGB(.01)", "XGB(.03)")
col_order <- c("Learner",
               "Structural / constrained",   "Structural / unconstrained",
               "First stage / constrained",  "First stage / unconstrained")

fmt <- function(b, se) sprintf("%.3f (%.3f)", b, se)

make_table <- function(df, instrument, ylab) {
  out <- df |>
    filter(instr == instrument, yvar == ylab) |>
    mutate(Learner = factor(learner_lbl[learner], levels = learner_order),
           Stage   = ifelse(spec == "structural", "Structural", "First stage"),
           Constr  = ifelse(constraint, "constrained", "unconstrained"),
           col     = paste(Stage, Constr, sep = " / "),
           cell    = fmt(estimate, std.error)) |>
    select(Learner, col, cell) |>
    pivot_wider(names_from = col, values_from = cell) |>
    arrange(Learner) |>
    select(any_of(col_order))
  # 2SLS doesn't depend on the constraint -- show value once per Stage
  # (in the constrained column) and put an em-dash in the duplicate.
  is_2sls <- as.character(out$Learner) == "2SLS"
  out[is_2sls, "Structural / unconstrained"]  <- "—"
  out[is_2sls, "First stage / unconstrained"] <- "—"
  out
}
```

# Machine Labor — DDML for the Angrist & Evans IV

The classical study of [Angrist & Evans (1998)](https://doi.org/10.1257/aer.88.3.450) estimates the effect of a third child on maternal labor supply using the instrument `samesex`, an indicator for the first two children sharing a sex. The rationale for using `samesex` is that same-sex parents are modestly more likely to have a third child, but — conditional on the sex of the first child and the sex of the second child — `samesex` is plausibly unrelated to the mother's labor market outcomes.

[Angrist & Frandsen (2022)](https://doi.org/10.1086/717933) revisit this same application in their article *Machine Labor* and ask whether ML can be used to estimate the first stage of two-stage least squares. 

They show that it can go badly: the random forests they consider discover the `boy1st * boy2nd` interaction inside the first-stage function $E[Z \mid X]$. As a consequence, the residualized instrument collapses to noise, and the second-stage coefficient becomes uninformative. They conclude that ML is "ill-suited" for instrumental variables in this application.

The identification assumption that makes 2SLS work is that `boy1st` and `boy2nd` enter additively, not interactively, in the first stage. This assumption is implied by the linear regression model and enforced by least squares. 

We show below that we can encode the additivity assumption into ML algorithms directly. With that restriction imposed, DDML and 2SLS yield similar estimates, and DDML also passes a placebo test.

## 1. Setup and variables

| Variable | Description |
| --- | --- |
| `workedm` | Outcome: mother worked last year |
| `morekids` ($D$) | Indicator: mother has more than two children |
| `samesex` ($Z$) | Instrument: first two children of the same sex |
| `boy1st`, `boy2nd` | Sex of first / second child (controls) |
| `agem1`, `agefstm` | Mother's age / age at first birth |
| `blackm`, `hispm`, `othracem` | Race indicators |
| `educm` | Years of education |

The script defines three named groups of regressors:

<details markdown="block">
<summary>R code — variable groups</summary>

```r
X_vars     <- c("boy1st", "boy2nd", "agem1", "agefstm",
                "blackm", "hispm", "othracem", "educm")
Xbase_boy1 <- c("boy1st",          "agem1", "agefstm",
                "blackm", "hispm", "othracem", "educm")  # drops boy2nd
Xbase_boy2 <- c(         "boy2nd", "agem1", "agefstm",
                "blackm", "hispm", "othracem", "educm")  # drops boy1st

D_var  <- "morekids"
Z_var  <- "instr" # either `same_sex` or a placebo instrument
kfolds <- 2L
```

</details>

`Xbase_boy1` and `Xbase_boy2` are deliberately set up so that we can build a polynomial dictionary separately within each block, guaranteeing the resulting features never multiply a `boy1st`-term by a `boy2nd`-term.

## 2. The identification issue, formally

The partially linear IV (PLIV) model is

$$
Y = \theta\, D + g(X) + \varepsilon, \qquad
D = m(X) + \alpha\, Z + u, \qquad
Z = r(X) + v.
$$

The DDML score uses the residualized instrument $\tilde Z = Z - \hat r(X)$ to instrument the residualized treatment $\tilde D = D - \hat m(X)$ in a regression on $\tilde Y = Y - \hat g(X)$. The whole construction collapses if $\hat r(X) \approx Z$, because then $\tilde Z \approx 0$.

In Angrist & Evans, by construction,

$$
\text{samesex} = \text{boy1st}\cdot\text{boy2nd} + (1-\text{boy1st})(1-\text{boy2nd}),
$$

which is *exactly* a function of the interaction `boy1st × boy2nd` and the levels. So if $r(X)$ is allowed to use that interaction, a flexible learner will recover `samesex` essentially perfectly from $X$ — making $\tilde Z = 0$ by construction.

OLS doesn't have this problem because, we never enter `boy1st × boy2nd` as a control. The exclusion of `boy1st × boy2nd` is an identifying restriction. Flexible machine learners like random forests and tree-based gradient boosting will leverage interactions for prediction and thus destroy the instrument. The cure is to enforce the same restriction inside the machine learning algorithms.

## 3. Two ways to encode the restriction

### 3.1 Polynomial dictionaries built block-by-block

For lasso and ridge we build a cubic polynomial dictionary *separately* on `Xbase_boy1` and `Xbase_boy2`, then stack them. Because the two dictionaries never share both `boy1st` and `boy2nd`, no resulting monomial multiplies the two sex indicators together.

<details markdown="block">
<summary>R code — `build_poly3()` and the block-structured design matrix</summary>

```r
build_poly3 <- function(df, base_cols) {
  X <- as.matrix(df[, base_cols, drop = FALSE])
  P <- poly(X, degree = 3L, raw = TRUE)
  out <- model.matrix(~ P)[, -1L, drop = FALSE]
  exps <- do.call(rbind, lapply(
    strsplit(sub("^P", "", colnames(out)), ".", fixed = TRUE),
    as.integer))
  colnames(out) <- apply(exps, 1L,
    function(e) paste(rep(base_cols, e), collapse = "_x_"))
  out[, order(rowSums(exps)), drop = FALSE]
}

X_poly_b1 <- build_poly3(data, Xbase_boy1)   # cubic poly without boy2nd
X_poly_b2 <- build_poly3(data, Xbase_boy2)   # cubic poly without boy1st
```

</details>

### 3.2 XGBoost interaction constraints

Tree-based methods allow imposing no-interaction constraints by enforcing that a single tree can split on `boy1st` *or* `boy2nd` but never both inside one path. Standard random forest implementations in R do, however, not implement such a constraint. We thus use XGBoost which lets us declare which features are allowed to appear together along any single tree path. We pass two groups, one containing `boy1st` (index 1) plus the demographics, and one containing `boy2nd` (index 2) plus the demographics. 

<details markdown="block">
<summary>R code — `make_learner_specs(constraint = TRUE, ...)`</summary>

```r
# X_vars order: 1=boy1st, 2=boy2nd, 3=agem1, 4=agefstm,
#               5=blackm, 6=hispm, 7=othracem, 8=educm
xgb_args1 <- list(nrounds = 500L, learning_rate = 0.01, nthread = 1L,
                  interaction_constraints = list(c(1, 3:8), c(2:8)))
xgb_args2 <- list(nrounds = 500L, learning_rate = 0.03, nthread = 1L,
                  interaction_constraints = list(c(1, 3:8), c(2:8)))

specs_stack <- list(
  list(what = ols,         assign_X = base_idx),
  list(what = mdl_glmnet,  args = list(alpha = 1, cv = TRUE),
                           assign_X = poly_idx),
  list(what = mdl_glmnet,  args = list(alpha = 0, cv = TRUE),
                           assign_X = poly_idx),
  list(what = mdl_xgboost, args = xgb_args1, assign_X = base_idx),
  list(what = mdl_xgboost, args = xgb_args2, assign_X = base_idx) 
)
```

</details>

The `assign_X` field tells `ddml` which columns of `X_full` each learner sees: OLS and XGBoost work on the eight original variables; lasso and ridge work on the (additivity-respecting) polynomial dictionary.

## 4. DDML estimation

We call `ddml::ddml_pliv()` with the five-learner stack, short-stacking via non-negative least squares, two folds, and heteroskedasticity-robust standard errors. After fitting, we also pull the per-learner cross-fitted residuals so we can read off a per-learner first-stage coefficient.

<details markdown="block">
<summary>R code — `ddml_pliv()` plus per-learner first-stage extraction</summary>

```r
fit_ddml <- ddml_pliv(
  y                       = y,
  D                       = D,            # morekids
  Z                       = Z,            # instr (one of three flavors)
  X                       = X_full,
  learners                = specs_stack,
  sample_folds            = kfolds,
  ensemble_type           = "nnls",
  custom_ensemble_weights = diag(length(specs_stack)),  # one "ensemble" per learner
  shortstack              = TRUE,
  silent                  = TRUE
)

# Per-learner first-stage: regress residualized D on residualized Z.
for (l in colnames(fit_ddml$fitted$D1_X$cf_fitted)) {
  D_resid <- D[,1] - fit_ddml$fitted$D1_X$cf_fitted[, l]
  Z_resid <- Z[,1] - fit_ddml$fitted$Z1_X$cf_fitted[, l]
  feols(D_resid ~ Z_resid, se = "hetero")   # learner-l first stage
}
```

</details>

Three things worth noting:

- `custom_ensemble_weights = diag(5)` makes `ddml` return *both* the NNLS stack and the five single-learner "ensembles," so a single fit yields all the per-learner numbers below.
- The PLIV score has the same Neyman-orthogonality structure as the partially linear model: small first-order errors in $\hat r$, $\hat m$, $\hat g$ wash out, *provided* $\tilde Z$ retains nontrivial variation. Section 3 is exactly about preserving that variation.

## 5. Two instrument flavors

To stress-test the pipeline we run it with two definitions of `Z`:

| Flavor | Definition | Expected behavior |
| --- | --- | --- |
| `observed` | `samesex` (the real Angrist–Evans IV) | DDML should agree with 2SLS |
| `fake` | `agem1 + educm + Uniform(0,1)` — a pure function of the controls plus noise | First stage and structural effect should both be ~0 |

`fake` is the cleanest placebo: any first-stage signal a learner finds there is spurious, because `Z` is by construction a function of two controls plus noise.

```{r run-loop}
#| echo: true
#| eval: false
df <- load_data("pums80m.dta")
res <- list()
for (flavor in c("observed", "fake")) {
  for (constraint in c(TRUE, FALSE)) {
    res[[paste(flavor, "workedm",
               ifelse(constraint, "constrained", "unconstrained"), sep = "_")]] <-
      estimate(df, yvar = "workedm", instrument = flavor, constraint = constraint)
  }
}
res <- bind_rows(res)
arrow::write_parquet(res, "results.parquet")
```

### 5.1 Real IV (`samesex`) — DDML reproduces 2SLS

For the observed `samesex` instrument, all constrained DDML variants — stacking, OLS, lasso, ridge, XGBoost — land on essentially the 2SLS point estimate. Unconstrained XGBoost (last two rows of the table) drifts off.

```{r res-observed-workedm}
#| echo: false
make_table(res, "observed", "workedm") |>
  kable(caption = "Outcome: workedm. Instrument: observed (samesex). Coefficient (SE) per learner; columns split Stage × Constraint. The 2SLS row does not depend on the constraint, so it is reported once per stage.")
```

### 5.2 Pure-noise IV (`fake`) — DDML refuses to find signal

When `Z = agem1 + educm + noise`, the true first stage is zero. Constrained DDML correctly returns a first-stage and structural effect indistinguishable from zero. Unconstrained learners — XGBoost in particular — can manufacture a nonzero first stage by exploiting the way `agem1` and `educm` enter both $r(X)$ and the controls.

```{r res-fake-workedm}
#| echo: false
make_table(res, "fake", "workedm") |>
  kable(caption = "Outcome: workedm. Instrument: fake (= agem1 + educm + uniform noise). The true effect is zero.")
```

## 6. Takeaway

Identification restrictions don't disappear when you replace OLS with a learner — they just move. In the Angrist & Evans first stage, the restriction is that `boy1st` and `boy2nd` enter additively, and it lives in our choice of regressors. Hand that choice off to a flexible ML learner and the restriction is gone; the learner will find the `boy1st × boy2nd` interaction, residualize the instrument away, and report an apparently-precise but uninformative second stage.

Two primitives make this restriction explicit and learner-side:

- **Block-structured feature dictionaries.** Build polynomial / interaction features within each block separately. Generalizes to "no interaction between A and B" for any partition.
- **`xgboost` `interaction_constraints`.** A list of allowed groups; any single tree path may only split on features that share a group. The same idea exists in some random-forest implementations under the name *additivity constraints* (see Hooker, 2007).

With either tool in place, DDML is well-behaved on the real IV *and* on the pure-noise placebo. ML is fine for the IV first stage — provided you tell the learner what it isn't allowed to learn.

## References

- Angrist, J. D., & Evans, W. N. (1998). Children and their parents' labor supply: Evidence from exogenous variation in family size. *American Economic Review*, 88(3), 450–477. <https://doi.org/10.1257/aer.88.3.450>
- Angrist, J. D., & Frandsen, B. (2022). Machine labor. *Journal of Labor Economics*, 40(S1), S97–S140. <https://doi.org/10.1086/717933>
- Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W., & Robins, J. (2018). Double/debiased machine learning for treatment and structural parameters. *Econometrics Journal*, 21(1), C1–C68. <https://doi.org/10.1111/ectj.12097>
- Ahrens, A., Hansen, C. B., Schaffer, M. E., & Wiemann, T. (2024). `ddml`: Double/debiased machine learning in R. *Journal of Statistical Software*, 108(3). <https://doi.org/10.18637/jss.v108.i03>
- Chen, T., & Guestrin, C. (2016). XGBoost: A scalable tree boosting system. *KDD '16*, 785–794. <https://doi.org/10.1145/2939672.2939785>
